---
title: "Lab 2: Exploratory Data Analysis"
author: "NAME"
format: html
embed-resources: true
html-math-method: mathml
execute: 
  warning: false
  echo: true
  error: false
---

## Before you start:

1. Complete the Lab Primer linked in Canvas  
2. Change the name of the **QMD** file only to your last name, first initial, or `OtterM-lab2-eda-quarto-student.qmd`. You will have to close the QMD document to rename the file.  
3. Be sure you have opened the `lab2-eda-student.Rproj` file.  
4. Change the "NAME" above after `author:` to your name, leaving it in quotes: "Monte Otter"  



## Skills

* Identify the population, sample, and variables of interest in a study
* Identify variable types
* Use R to generate and interpret descriptive statistics and graphics for numeric variables
* Editing graphics in R using arguments
* Learning and applying Markdown language for text editing
* Learning and applying Quarto documents for R code and output generation


These skills correspond to LTs 1, 2, and 3 and R LTs R1 & R2. 



**Reminder**: Before starting this lab, let us load the `tidyverse`, `ggformula` and `mosaic` packages we need for statistical and graphical functions.  

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(ggformula)
library(_packageforstats_) # add the package name we use for summary statistics 
```



***

## Introduction

In this lab, we're going to be learning both Exploratory Data Analysis for single quantitative (numeric) variables, and also Quarto. Quarto allows us to create documents with embedded code and outputs. You should read the accompanying lab Primer for a guide to Quarto.


***

## Part 1: Data Collection: Basal Metabolic Rate

As scientists, we need to not only be able to identify the research question, population, sample, and variables of interest for our own questions, but for other studies as well. The following is an abstract of a recent publication. Read it carefully, and then answer the questions about it.

#### Abstract

Basal metabolic rate (BMR) is posited to be a fundamental control on the structure and dynamics of ecological networks, influencing organism resource use and rates of senescence. Differences in the maintenance energy requirements of individual species therefore potentially predict extinction likelihood. If validated, this would comprise an important link between organismic ecology and macroevolutionary dynamics. 

To test this hypothesis, the BMRs of organisms within fossil species were determined using body size and temperature data, and considered in the light of species' survival and extinction through time. Our analysis focused on the high-resolution record of Pliocene to recent mollusks (bivalves and gastropods) from the Western Atlantic. Species-specific BMRs were calculated by measuring the size range of specimens from museum collections, determining ocean temperature using the HadCM3 global climate model, and deriving values based on relevant equations. Intriguingly, a statistically discernible difference in metabolic rate exists between those bivalve and gastropod taxa that went extinct and those that survived throughout the course of the Neogene. This indicates that there is a scaling up from organismic properties to species survival for these communities. Metabolic rate could therefore represent an important metric for predicting future extinction patterns, with changes in global climate potentially affecting the lifespan of individuals, ultimately leading to the extinction of the species they are contained within. We also find that, at the assemblage level, there are no discernible differences in metabolic rates for different time intervals throughout the entire study period. This may suggest that Neogene mollusk communities have remained energetically stable, despite many extinctions.

Indicate the correct answer by bolding your chosen response(s) around the answer, but not including the letter, for example: a. **this answer**.   

### 1.1 

What is the research question in this study? 

  a. What is the metabolic energy rate of mollusks?  
  b. Can differences in metabolic requirements predict extinction likelihood?   
  c. Does basal metabolic rate influences resource use?  


### 1.2

What is the population about which the researchers want to make inferences?  

  a. All organisms in the fossil record  
  b. All bivalves and gastropods  
  c. Pliocene to recent bivalves and gastropods  
  d. Pliocene to recent mollusks from the Western Atlantic  
  
  
### 1.3 

What is the sample that the researchers collected to make inferences about the population?  

  a. Bivalves and gastropods  
  b. Pliocene to recent bivalves and gastropods  
  c. Pliocene to recent mollusks from museum collections  
  d. Organisms in the fossil record  


### 1.4

What is/are the variable(s) that the researchers *directly* measured for each individual? (choose all that apply)  
  a. Basal metabolic rate (BMR)  
  b. Ocean Temperature  
  c. Specimen Size  


### 1.5

What type of study was this?    
  a. Observational   
  b. Experimental  




***

## Part 2: Curry

Stephen Curry is one of the most prolific scorers currently in the NBA. We are interested in the number of points he scores per game across his career. The data of the number of points Stephen Curry scored during each regular season game from 2024-2025 can be found in `curry2024.csv`.  


### 2.1

Modify the code below to read in the data `curry2024.csv` and store it as the object `curry` by replacing the blanks (some hints provided as: `___hint___`, or blanks simply as  ______).

```{r}
#| label: curry-data
curry <- ___goal___("___file___")
```


### 2.2

We might want to know more about the data that is inside `curry`. We can use the `names()` function to display the variable names to use in your code, while `str()` will display the type of variable R thinks the variables are (numeric: "int", or categorical: "Factor") plus the first few observations in the vector. Modify the code by supplying the dataframe object name to both codes in place of the blanks to inspect your variables. List each variable of interest (only) below and indicate its full variable type (e.g. categorical nominal), based on what you see in the code output. You should provide a justification for the variable type.  

```{r}
#| label: curry-names
names(______)
str(______)
```

> Variable: Type -- Justification  


### 2.3 

What is the population and sample for this scenario?

> Insert answer & justification here.  
 


### 2.4 

Create a histogram of the number of points Curry scores in a game by modifying the code below. Notice that this code is spaced out differently than we've seen before. It is just to make commenting easier -- R will keep reading the function until it reaches the end `)`. Comment on each line of code to indicate what it does and modify the code to add labels to the graph axes. **Don't forget good graphic guidelines when labeling your axes!**

```{r}
#| label: curry-hist1

gf_histogram(___formula___, # comment here
             data = ______, # comment here
             xlab = "x label here", # comment here
             ylab = "y label here", # comment here
             color = "black") # comment here

```


### 2.5

Reproduce your code from 2.4 below to create a histogram again, but this time include an additional argument, `binwidth = 5`. What changed in your graphic? 

```{r}
#| label: curry-hist2


```

> Insert answer here.



### 2.6 

We want our graphic to be easy to read, including the bin ranges. Reproduce your code from 2.5 below to create another histogram, but this time include an additional argument: `boundary = 0`. What changed in your graphic?

```{r}
#| label: curry-hist3


```

> Insert answer here.


### 2.7 

Create a boxplot of the number of points Curry scores in a game by filling in arguments to the function below. Do not add additional arguments other than the three asked for here.

```{r}
#| label: curry-boxplot

gf_boxplot(~ ______, 
           data = ______,
           xlab="x axis label here")
```



### 2.8

Describe the distribution of the number of points Curry scores in a game, using both the histogram and boxplot, being sure to address all four characteristics of a distribution (see your notebook, page 2-4). 

> Insert answer here.


### 2.9

Based on your description, what are the most *appropriate* measures of center and spread for these data? Justify your answer. 

> Insert answer here. 


### 2.10 

How many games are in our sample? Add code below to determine the `length()` of your variable (the number of cases/rows in the dataset). `length()` requires the `dataframe$variable` input, not the formula (~) template.

```{r}
#| label: curry-length

```

> Insert your answer here.


### 2.11 

We can use `length()` to find the number of cases in our dataset, but we know we aren't actually graphing that many data points, as Curry didn't play every game. Use `df_stats()` to determine the number of games that Curry did and didn't play. How many games did and didn't Curry play?

```{r}
#| label: curry-n

```

> Insert answers here.


### 2.12 

Calculate the appropriate summary statistics for the number of points Curry scores in a game. Modify the code (and replicate as needed) to calculate the values. Write your answers below using mathematical notation, for example, using a random number: $s = 10$. 

**Remember you need to load the `mosaic` package, in the `setup` code at the beginning of the document, first before we calculate these statistics.**  

```{r}
#| label: curry-stats

__goal__(~ ___variable___, data = ___dataframe___)
< add additional functions as needed > 
```

> Insert answers here.
> (remember to put two spaces after each line to force a line break.) 


### 2.13 

Using in-line code, write a sentence that describes the center and spread of the dataset. You may need to insert a code chunk here, or you may not, depending on how you choose to perform this task. 

> Insert your answer here.



### 2.14. 

Modify the code below to use the function that will produce the five number summary on which a boxplot is based. 

```{r}
#| label: curry-5num

______(~ ______, data = ______, <additional argument if needed>)
```


### 2.15

Compare your two measures of center. What can you determine about the shape of the distribution? Be sure to include a justification.  

> Insert your answer here. 


 


***

## Part 3: Winter Olympic Games  

The Winter Olympic games will soon be fascinating the world. While watching, Jordyn became interested about the number of medals won by countries across all Winter Olympics, but also noticing that the games has grown from just including 16 participating nations in 1924 to 91 nations in 2022. They are interested in looking at the medal count per involvement in the games. In their sampling, they allowed the same chance for each country to be sampled from all participating countries in Winter Olympic games from 1924 - 2022. They recorded the number of total medals a country won, the number of games the nation participated in, and calculated the ratio of medals per games in `WinterOlmypics_S26.csv`. The variable inside is `Ratio.MedalPart`.  



### 3.1 

What is the population and sample for this scenario?

> Provide an answer and justification here.  


### 3.2

Describe and name the sampling method used to select the sample. 

> Provide an answer and justification here.  


### 3.3 

What is the variable of interest recorded in `Ratio.MedalPart` and its type (categorical nominal, etc.)?  

> Provide an answer and justification here.   


### 3.4 

Insert a code chunk that will read in your data and add additional code chunks to create an appropriate graphic of your sample data. It should include any modifier as necessary to label the graphic.  

**Add code chunk here.**


### 3.5

Calculate the most relevant summary statistics. After your code chunk, describe each statistic using mathematical (i.e. $\bar{x} = 1$) and verbal notation (i.e. in a sentence with context). 

**Add code chunk here.**

> Provide mathematical (e.g. $equation$) and verbal interpretation of each value calculated. 


### 3.6

Describe the shape of your distribution. Based on the shape of the distribution, how do the mean and median values relate to each other? Justify your answer.  

> Describe the distribution's shape.  
> How do mean and median values relate to each other & why.  


### 3.7

What percentage of countries have a medal ratio of more than 1 medal per games?   

> Insert answer here.  


### 3.8

Do more countries have a medal ratio below 0 compared to those with a medal ratio above 0? Explain your answer.   

> Insert answer here.  


### 3.9 

Look at the graphic you created in 3,4. What information do you learn about the winter Olympic medal ratio of countries? How do your measures of center agree with the information you learned from the graphic?

> Insert answer here. 



Before you finish -- check your rendered document. Is it easy to read? Do you need to add spaces anywhere to create line breaks? Did you accidentally delete any of the answer formatting?




