---
title: 'Lab 8: Analysis of Variance'
author: "NAME"
format: html
embed-resources: true
html-math-method: mathml
execute: 
  warning: false
  echo: true
  error: false
---


## Before you start:

1. Complete the Lab Primer linked in Canvas.  
2. Be sure you have opened the `lab8-anova-student.Rproj` file.  
3. Change the name of the **QMD** file only to your last name, first initial, e.g., `OtterM-lab8-anova-student.qmd`. You will have to close the QMD document to rename the file.   
4. Change the "NAME" above after `author:` to your name, leaving it in quotes: "Monte Otter"  


## Skills

*	Explain the purpose of a multiple comparison procedure to compare multiple means.
* Be able to identify a situation that requires ANOVA.  
    + Identify the factor (i.e. explanatory variable) and response variable.  
    + Be able to state the appropriate hypotheses for ANOVA.  
    + Report ANOVA results in an ANOVA table and be able to interpret each part of the table within the context of the study.  
* Explain the necessary conditions to use an ANOVA  
    + Be able to assess the conditions.  
* Be able to identify a situation that requires an ANOVA for transformed data
    + Identify the factor (i.e. explanatory variable) and response variable
    + Be able to state the appropriate verbal-only hypotheses for a transformed ANOVA
    + Report ANOVA results in an ANOVA table and be able to interpret each part of the table within the context of the study.
* Be able to interpret output from a post hoc test (Tukey's method).
* Be able to explain the difference between multiple t-tests and a post hoc test.
* Be able to execute the proper R codes to conduct ANOVA and post hoc analyses.

The above skills cover LT 1-5, 10-16, 21c, 22ab, and 23a.  It also introduces R-LT R7. 


Here are some symbols that you might find useful:  
$H_0$ $H_1$ $\sigma$ $\mu$ $\mu_0$ $\mu_{word}$ $\mu_{d~word}$ $\neq$


Update the following code chunk with the appropriate packages.  

```{r}
#| label: setup
#| include: false
library(tidyverse)

```



***

## Scenario 1: Penguin Diets      

Data on the diets and foraging strategies of Subantarctic penguins during the winter, when individuals migrate far from their breeding grounds and field work is difficult, is lacking. Prior research suggests that winter diets are more variable as penguins migrate, but the summer diets around the breeding grounds are dominated by one fish species. The researchers plan to analyze both the winter prey composition through an ANOVA of stable isotope data and to analyze the summer prey composition through an ANOVA of prey length. Let's examine the prey length data. 

Researchers were able to study the length of the current prey -- myctophid fish -- for the adults of three of the penguin species present on Crozet Island: King penguins, Macaroni penguins and Rockhopper penguins. The data for the prey standard lengths, in mm, collected from a random sample of each penguin species, is found in `prey.csv`. 

**Researchers want to determine if there is a difference in the standard length of prey (myctophid fish) eaten by the three different penguin species.**


<Insert code chunk here and read in the data> 



### 1.1.  

Since the researchers plan to make comparisons among the three species of penguins, they chose first to conduct an ANOVA test on the prey length based on species before making comparisons between each species of penguin. Why is it better to conduct an ANOVA first before making comparisons between each pair of penguin species?  

> Insert answer here   


### 1.2.  

Name the response and explanatory variables of interest and, for each, state the type of variable (quantitative or qualitative; continuous, discrete, nominal, or ordinal). Recall this research question is examining the effect of penguin species on eaten myctophid fish length.  

> Response: Insert answer here     

> Explanatory: Insert answer here   


### 1.3.  

State the null and alternative hypotheses, symbolically and verbally as appropriate.

> Insert answer here   


### 1.4.  

Modify the code below to conduct an appropriate testing of the conditions to calculate an ANOVA analysis. State each condition, state the evidence you used to assess the condition (e.g. graphic), and then assess the evidence for if the condition is met or not. Use the template below the code chunk for your answer to each condition.  

```{r}
#| label: condition-checks

species_aov <- [insert code that does an anova here]

<insert code that creates a qq plot of residuals here>

<insert code that creates a residuals plot here>
```

> Condition #_: ___     
> Evidence: ___    
> Justification: ___    

> insert more condition templates as needed  



### 1.5.  

Insert a code chunk and write the code below to print out an anova table for your analysis. Fill in the values of the appropriate statistics.

<Insert code chunk here>   

> df~e~:    
> df~t~:    
> MSE:    
> MST:     
> F:    
> p-value:   


### 1.6.  

Interpret your p-value in the context of the scenario.

> Insert answer here   


### 1.7.  

Write a strength of evidence statement for your p-value, using the significance level $\alpha = 0.025$.  

> Insert answer here   


### 1.8.  

Based on your exploratory data analysis (EDA), what would be your conclusion in answer to the research question about if there are differences in eaten myctophic fish length among the three different penguin species? You will need to insert a code chunk to calculate the appropriate statistics and create a graphic to answer this question. Be sure to cite your evidence and justify how it supports the conclusions about the parameters.   

```{r}
#| label: exploratory-data-analysis
<insert summary statistics code here>
<insert boxplot code here>
```

> Insert answer here   



### 1.9.  

Is there any reason to do a Tukey test on the prey lengths? **Bold** the words (but not the list-letter) of your answer.   

a. No, we risk elevated Type I error with multiple comparisons in a Tukey test.  
b. No, there is no reason to need a comparison.  
c. Yes, we should do a Tukey test to test our original null hypothesis.   
d. Yes, we should do a Tukey test to identify which pairs are significantly different.  


### 1.10.  

What is the consequence of using an independent two-sample t-test for each pairwise comparison (instead of using the Tukey test)? **Bold** the words (but not the list-letter) of your answer.  

a. We risk an increased Type I error rate in each test.  
b. We risk an increased chance of committing a Type I error.  
c. We risk an increased chance of committing a Type II error.  
d. We risk eliminating random coincidences as a reasonable explanation.  


### 1.11.  

Conduct a Tukey test by modifying the code below. (1) Fill out the table below to organize the results and (2) then provide a written summary of your *post hoc* test results of all possible factor pairings. In your written summary, you should address all pairwise comparisons in terms of if they are the same, if they are different (i.e., at least moderately strong evidence against the null, using the family-wise significance of 0.05, and include p-values as evidence), and, if it exists, the direction of difference. *This is not a list. Simply state generally how parameters compare, the way you might communicate verbally with a scientist, with evidence.*

```{r}
#| label: tukey-comparisons
______(species_aov)
```


  pair   | p-value | strong evidence? (Y/N) | how do they compare? (use <,>,or =)
---------|---------|------------------------|---------------------------------------
  M-K    |         |                        |    
  R-K    |         |                        |    
  R-M    |         |                        |  


> Insert answer here  


 
### 1.12. 

It has been proposed that so many species of penguins can coexist on the Crozet Islands because they partition the food supply (e.g. eat different prey items, or different sizes of the same prey item). Does your analysis support or refute this proposal? Explain if so or if not, and any possible reasons.      

> Insert answer here   


### 1.13.  

Do you trust the results of your ANOVA and Tukey tests? Explain why with statistical reasoning. If you do not trust the results, suggest and justify an alternate analysis. 

> Insert answer here   


### 1.14. 

Given our scenario, **bold** any answers (but not the list-letter) below that is a true statement.  

a. The penguins were sampled randomly, so this can only be an observational study and thus we cannot determine causality.
b. The penguins were sampled randomly, so this can only be an observational study and thus we cannot generalize to all the penguins in each species.
c. The penguins were not randomly assigned to each treatment group (species), so this is an observational study and thus we can generalize to all the penguins in each species.  
d. The penguins were not randomly assigned to each treatment group (species), so this is an observational study and thus we cannot determine causality.   




***



### Scenario 2: Iron in Chesapeake Bay Waters       


The Chesapeake Bay is a very important feature for the ecology and economy of the states of Maryland and Virginia. During the Spring, light fresh-water from rivers flows into Chesapeake Bay and down toward the ocean while saltier, denser water flows up into the bay from the ocean. The difference in density means these water bodies don't mix, but instead form stratified layers in the water column. The lower layer of salty water has a lower pH and oxygen deficiency. Unfortunately, researchers think that layer also is the source of the smell of hydrogen sulfide from the sulfate-reducing bacteria, *Desulfovibrio*, which can live in low- to no- oxygen environments. Since iron is known to be vital in the nutrition of *Desulfovibrio*, researchers wanted to know if there is a difference in the amount of iron present in different depths of the Chesapeake Bay. To answer this, the researchers randomly selected different locations of the Chesapeake Bay to gather samples of the water at 0ft, 25ft, 50ft and 100ft and measured the amount of iron (mg/L) present. The data are `iron-chesapeake.csv`. 

**Researchers want to determine if the iron level of the waters of Chesapeake Bay is different across the four depths**.


### 2.1. 

State the null and alternative hypotheses, symbolically and verbally as appropriate.

> Insert answer here   


### 2.2. 

State each condition, state the evidence you used to assess the condition, and then assess the evidence for if the condition is met or not. Be sure to fix your explanatory variable for an ANOVA analysis.     


<Insert code chunk here>   

> Insert answer here   



*** 
If the parametric conditions of ANOVA are not met, we can consider using a transformation before using a non-parametric option. In your class notes, we gave you some general guidelines for choosing possible transformations to try on your data. We will perform a logarithmic transformation on the data from each location using the `log` function. 


### 2.3. 

*Challenge Question* What do you notice about the distribution of iron levels in Chesapeake Bay that might suggest a logarithmic transformation would be appropriate? *(Hint: a boxplot, calculating summary statistics, or looking back at your condition plots might help you determine this)*

<Insert code chunk here>  

> Insert answer here   

 

### 2.4. 

Name the explanatory and the response variables of interest for the **logarithmic-transformed** ANOVA and state the type of variable (quantitative or qualitative; continuous, discrete, nominal, or ordinal). 

> Response: Insert answer here    
> Explanatory: Insert answer here     



### 2.5.   

State the null and alternative hypotheses for the **logarithmic-transformed** ANOVA, verbally only.

> Insert answer here   


### 2.6.   

Assess all the conditions of the **logarithmic-transformed** ANOVA. State each condition to conduct the inference test, state the evidence you used to assess the condition, and then assess the evidence for if the condition is met or not.    


```{r}
#| label: log-aov

log_iron_aov <- <insert function>(log(<insert reponse>) ~ factor(<insert explanatory>), data = <insert dataname>)


```


> Insert answer here   



### 2.7. 

Insert a code chunk that will print out your ANOVA table from the *logarithmic-transformed* analysis. Fill in the appropriate statistics below. 

<Insert code chunk here>   


> df~e~:     
> df~t~:     
> MSE:     
> MST:      
> F:     
> p-value:     



### 2.8.

Write a statement of the strength of your evidence of your p-value from the **logarithmic-transformed** ANOVA.   

> Insert answer here   


### 2.9. 

If the ANOVA on transformed data yields significant differences between the four depths, we will want to conduct a *post-hoc* analysis. Have we met the conditions to perform a post-hoc Tukey test? (yes or no & why; do not re-assess).

> Insert answer here   


### 2.10. 

Insert a code chunk to conduct a Tukey test. Provide a written summary of your *post hoc* test results of all possible depth pairings. 

< Insert code chunk here>    

> Insert answer here   



### 2.11.   

Given our scenario, **bold** any answers (but not the list-letter) below that is a true statement.

a. The cases were not randomly assigned to each treatment group, so this is an observational study.    
b. The cases were sampled randomly, so this can only be an observational study and thus we cannot determine causality.   
c. The cases were not sampled randomly, so we cannot generalize to all Chesapeake water within the three species.   
d. The cases were sampled randomly, so we can generalize our results to all cases at these depths.   
e. The cases were randomly assigned to each treatment group, so this is an experiment, so we can say any changes we observed is due to the Chesapeake Bay depths   
f. The cases were randomly assigned to each treatment group, so this is an experiment, but we cannot determine the cause of the difference.  




