---
title: 'Lab 6: Introduction to Hypothesis Testing (with Two Independent Groups)'
author: "NAME"
format: html
embed-resources: true
html-math-method: mathml
execute: 
  warning: false
  echo: true
  error: false
---


## Before you start:

1. Complete the Lab Primer linked in Canvas.  
2. Be sure you have opened the `lab6-hyp-test-student.Rproj` file.  
3. Change the name of the **QMD** file only to your last name, first initial, e.g., `OtterM-lab6-hyp-test-student.qmd`. You will have to close the QMD document to rename the file. 
4. Change the "NAME" above after `author:` to your name, leaving it in quotes: "Monte Otter"



## Skills

* Given a dataset, be able to determine if the appropriate assumptions are met to conduct a parametric hypothesis test using probability plots (qq plots) to assess the assumption of normality and exploratory data analysis to assess the guideline of equal variance  
* Explain that in order to determine if there is a relationship or underlying circumstance that explains the observed characteristics of our sample data, we must compare that observation of a perceived difference or relationship to what might occur simply by random chance.
* Identify and define the null and alternative hypothesis.  
* Evaluate p-values, test statistics and other information to assess the results of a hypothesis test.  
* Interpret a p-value in context
* Draw specific and general conclusions based on the results of the hypothesis test; interpret the information in context.
* Conduct statistical inference (confidence interval or hypothesis test) based on two independent samples   
* Utilize R-code to conduct two-sample statistical inferences (confidence intervals, hypothesis tests)   


These skills correspond to LTs 14, 15, 21a, and 22b and R LT R5.


***

Here are some symbols that you might find useful:  
$H_0$ $H_1$ $\sigma$ $\mu$ $\mu_{word}$ $\neq$


Before you start the lab, add the code that will load any other packages you might need into this code chunk. 

```{r}
#| label: setup
#package for data management
#package for graphics
#package for stats

```


## Part 1: Chicken Weights

Chicken farming is a multi-billion dollar industry, and any method that increases the growth rate of young chicks can reduce consumer costs while increasing company profits, possibly by millions of dollars. A study was conducted to measure and compare the effectiveness of feed supplements on the growth rate of chickens. A chicken farm in central California selected every n-th chick, where n is a random number between 1 and 10. Those chicks were randomly allocated into two groups, one of which was fed a horsebean supplement and the other a linseed supplement. After six weeks, their weights in grams were recorded. The researchers are interested in determining which feed type is more effective at increasing weight and therefore growth rate of chickens. The data are stored in `chickwts.csv`.


### 1.1.  

State the parameter(s) of interest.

> [parameter1 symbol] = [parameter of interest in words]   
> [parameter2 symbol] = [parameter of interest in words]  


### 1.2. 

State the study type of this study (observational or experiment). Justify your response.

> Insert answer here.   
> Insert justification here.  

 
### 1.3.   

State your null and alternative hypotheses, both symbolically and verbally. You may need one of these symbols for the alternative: $<, >, \neq$.

> $H_0$: [insert symbolic hypothesis here]  
> Insert verbal hypothesis here.  

> $H_1$: [insert symbolic hypothesis here]  
> Insert verbal hypothesis here.  


### 1.4. 

Check the conditions to use the Student's t distribution to conduct the hypothesis test. The conditions are listed for you; answer if they were met and justify why in the context of the question. Include any code needed to assess the conditions.

```{r}
#| label: read-chicks
chick <- read_csv("chickwts.csv", show_col_types = FALSE)
```

```{r}
#| label: check-conditions-chicks

[insert code to check condition 2a]

[insert code to check condition 2b]

```


> 1. Random sample: [Met?]; [Justification]  
  2. Normality or Sufficient sample size:  [Met?]  
      2a. Is the sample from a Normally distributed population or does the sample data have a roughly bell-shaped or symmetric distribution? [Yes / No]; [Justification]       
      2b. Is the sample size sufficient? [Yes / No]; [Justification]    
      
> Overall decision: [Met?]   



### 1.5. 

Calculate the appropriate test statistic and p-value 'by hand' (i.e., using the numerical summary functions in R, without using the `t.test()` function); show your work. Provide the test statistic and p-value below.

```{r}
#| label: calc-t-test-chicks

se <- [insert code to calculate standard error here]

t <- [insert code to calculate test statistic here]

df <- __n.smaller__ - 1

## To calculate p-value, based on alternative symbol - Choose the correct ONE code below based on the context; delete the others so that you only leave the one correct code to run.

# <
pt(t,df)

# >
1-pt(t, df)

# =/= and t < 0
2*pt(t, df)

# =/= and t > 0
2*(1-pt(t, df))
```

> Record the test statistic and p-value that is returned  
> t = ____, df = ____, p-value = ____


### 1.6. 

Interpret your p-value in the context of the scenario. 

> There is a [insert p-value] probability of getting our test statistic of [symbol = insert test statistic] with degrees of freedom of [insert degrees of freedom] or [insert direction], assuming [insert null hypothesis].   


### 1.7.  

Provide a strength of evidence statement from your p-value, in the context of the scenario, using a significance level of 0.05.    

> Insert answer here   


### 1.8. 

Calculate the 95% confidence interval (CI) for the parameter of interest. Calculate, using numerical summary functions in R, as well as one line of code to determine the critical value (denoted by t*). Remember that you've already calculated your standard error in part 1.5.

```{r}
#| label: calc-ci-chicks
pe <- [insert code to calculate point estimate here]

cv <- [insert code to calculate t* here]

# Calculate lower bound and upper bound using previously stored values (remember you alreaqdy calculated some values previously)
(lb_wts <- _____ - _____*____)
(ub_wts <- _____ + _____*____)
```


### 1.9. 

Provide a second piece of evidence from EDA (Exploratory Data Analysis) that supports the results of the hypothesis test in determining which feed type is more effective at increasing weight of chicks. Be sure to cite your evidence and make a conclusion about the parameter.     

< Insert code chunk here > 

> Insert answer here   


### 1.10. 

Identify the population, if any, to which the study can be generalized. Explain why with evidence.     

> Insert answer here.  



***

## Part 2: Dolphin Blubber Insulation    

For marine mammals whose young are born directly into the ocean, that comes with a unique challenge. The ocean is a fairly cold environment, and water has a high heat conductivity, meaning it transfers temperatures easily. This means that thermal regulation (retaining body heat) is especially important for newborn whales and dolphins in this cold water. Researchers know that in Atlantic bottlenose dolphins *Tursiops truncatus*, both the thickness and quality (lipid content) of blubber varies across different life stages and influences insulation. One measure of the *quality* of blubber is its "conductance," measured in watts per square meter per degree Celsius. Lower conductance scores means higher quality and insulation, and thus provides better protection from the cold. The researchers are interested if Atlantic bottlenose newborn dolphins are better protected from the cold compared to adults. They measure conductance of blubber in adult and newborn Atlantic bottlenose dolphins. For the order of the difference, use `Newborn - Adult`.
*Based on Dunkin et al. 2005, J. Exp. Biol.*   



### 2.1.  

State the parameter(s) of interest in the study.

> Insert answer here  


### 2.2.  

State your null and alternative hypotheses, both symbolically and verbally.

> Insert answer here  


### 2.3.  

Read in the data from your csv file `blubber.csv` and use `mutate()` to re-order the levels of the categorical explanatory variable to set the order of the difference `Newborn - Adult`.  

< Insert code chunk here >  


### 2.4.  

Create a boxplot of your sample from each group in one single graphic (i.e., numerical variable by categorical variable). Be sure to label the axes appropriately and fully. 

< Insert code chunk here > 


### 2.5.  

State and assess the conditions to use the Student's t distribution to conduct the hypothesis test in the context of the problem. Use the skills you learned in Lab 5 as needed to complete this question. Include any code needed to assess this problem. Be sure to provide full justifications.  

<Insert code chunk here>

> 1. [Condition 1]: [Met?]; [Justification]  
  2. [Condition 2]:  [Met?]  
     2a. [evidence and justification in context of question]   
     2b. [evidence and justification in context of question]    
     
> Overall decision: [Met?]   



### 2.6.  

Perform an appropriate hypothesis test using the `t.test()` function, then provide your test statistic, degrees of freedom (Welch's approximation), and p-value.

<Insert code chunk here>

> Insert answer here  


### 2.7.  

Interpret your p-value in the context of the scenario.

> Insert answer here  


### 2.8.  

Provide a strength of evidence statement from your p-value, in the context of the scenario, using a significance level of 0.01.   

> Insert answer here  


### 2.9.  

Using the `t.test()` function, calculate the 99% confidence interval for the difference in population means of blubber conductance of newborn Atlantic bottlenose dolphins and of adult Atlantic bottlenose dolphins.

<Insert code chunk here>


### 2.10.  

Based on your confidence interval, is there evidence that newborn dolphins are better protected from the cold compared to adults? What can you conclude about the blubber conductance (watts per square meter per degree Celsius) of newborn Atlantic bottlenose dolphins compared to adult dolphins? Evaluate the confidence interval for evidence and justify your answer in the context of the parameter of interest and the research question.    

> Insert answer here  


### 2.11.  

Can the researchers determine if the life stage of Atlantic bottlenose dolphins *caused* any differences in the blubber conductance? Justify your answer fully.   

> Insert answer here  






