---
title: "Lab 10: Simple Linear Regression"
author: "NAME"
format: html
embed-resources: true
html-math-method: mathml
execute: 
  warning: false
  echo: true
  error: false
---


## Before you start:

1. Complete the Lab Primer linked in Canvas.  
2. Be sure you have opened the `lab10-regression-student.Rproj` file.  
3. Change the name of the **QMD** file only to your last name, first initial, e.g., `OtterM-lab10-regression-student.qmd`. You will have to close the QMD document to rename the file. 
4. Change the "NAME" above after `author:` to your name, leaving it in quotes: "Monte Otter" 


## Skills

* Given a study, distinguish between explanatory and response variables
* Given a set of raw data, make a scatterplot or regression output using R
* Given a scatterplot, identify patterns such as positive and negative associations
* Given the least squares line and a value of x, calculate the predicted value of y
* Given standard regression output, identify and utilize key parts of the output (estimated slope, intercept, R^2^, etc.)
* State and test the assumptions of inference about the regression model
* Given a study objective, significance level, and summary statistics, conduct a formal test of significance on a slope based on ANOVA or the t-distribution by conducting the appropriate steps (including stating hypotheses, calculating test statistics, calculating and interpreting p-values and interpreting the conclusion in context).
* Be able to calculate and interpret the confidence interval for $\beta$ given output from a linear regression analysis. 
* Given standard regression output, interpret the results of the test of hypothesis about the slope.
* Given a study, interpret the value of the coefficient of determination (R^2^).
* Be able to describe the primary distinction between regression and correlation analysis.

New LTs covered in this lab include 17, 18, 20b, 23abc, plus 22ab, and others. R LTs include R2, R8, and R9.  


Here are some symbols that you might find useful:  
$H_0$  $H_1$  $\neq$  $\sigma$  $\mu$  $\mu_0$  $\mu_{word}$  $\rho$  $\beta_0$  $\beta_1$  $\hat{y}$  $\hat{y}_{word}$  $value \cdot x_{word}$  
 
***
*Markdown tip: putting two spaces after a line forces the next line in your markdown document to also be a new line in your knitted document.*

*Markdown tip: equations in Markdown will ignore any attempt to include spaces, unless you insert a `~` inside the equation where you want the space.* For example, in our BAC-Beers problem from class (hover your mouse over them to see what it would knit like): $x_{number of beers}$ versus $x_{number~of~beers}$ 

*Markdown tip: If you want to test out two lines of code, such as piped functions, you have to run them __together__, not one at a time.* 

*Markdown tip: you can run one line of code at a time by placing your cursor in it, then Ctl + Enter*
***

**Reminder:** Before you start the lab, add the code that will load any other packages you might need into this code chunk. 

```{r}
#| label: setup
#| include: false
library(tidyverse)

```


## Scenario 1: Brain Weight in Healthy Adults   

A study published in 1905 had the goal of predicting the approximate weight of the human brain in healthy adults and theorized that the size of the head would be a good predictor. The researchers at Middlesex teaching Hospital decided to focus on post-mortem patients who had suffered accidental deaths, as a reasonable way to collect the data for the population of healthy adults. Over the period of three years, they randomly selected post-mortem patients and for each case, measured first their heads (cm^3^) and then took out the brain and weighed it (in grams). Here the size of the head is defined as the product of the three principal diameters of the head. Additionally, the patients’ sex (male or female) and age group (20-46 or 46+) was recorded. The researchers start by completing a linear regression analysis for the brain weight against head size in all healthy adults; they expect there may be a positive linear relationship, as that makes sense given that the brain is contained in the head. The data is in `brain-head.csv`.    
*Source: R.J. Gladstone (1905). "A Study of the Relations of the Brain to to the Size of the Head", Biometrika, Vol. 4, pp105-123*   
 

### 1.1.

Identify the variable type by completing the table below. Please **bold** your answers.

Variable       | Description                            | Type (Categorical or Numeric)
---------------|----------------------------------------|-------------------------------
BrainWeight    | Brain weight (grams)                   |     
HeadSize       | Head size (cm^3^)                      |     
Sex            | Adults' biological sex (male / female) |    


The researchers start by completing a linear regression analysis for the brain weight and head size in all healthy adults; they expect there may be a positive linear relationship, as that makes sense given that the brain is contained in the head.    


### 1.2. 

State the null and alternative hypotheses for a regression hypothesis test of brain weight and head size in healthy adults, both symbolically and verbally.  

> Insert your symbolic null here   
> Insert your verbal null here  

> Insert your symbolic alternative here  
> Insert your verbal alternative here  



### 1.3. 

Suppose the parametric conditions are met for this analysis. Conduct a linear regression analysis on these data. Modify the code below so it will create a linear model and display the summary table.  

```{r}
#| label: build-model-full

braindat <- read_csv("brain-head.csv", show_col_types = FALSE) # do not change this code

brain.lm <- lm(_____ ~ _____, data = ______)
summary(_____)
```


### 1.4. 

Does the linear model with head size explain much of the variability in brain weight in healthy adults? How much variability does it explain? Does that seem like a good model to use for predictions (evaluate the strength of the model fit)?    

> Insert answer here  



The researchers notice that the strength of linear relationship may depend on sex of the person. To this end, the researchers considered two separate linear models, one for males and the other for females: 

```{r}
#| label: scatterplot-sex
## this graphic uses the same structure that you have learned, just color-codes given the cases' categorical variable state (color = ~Sex). The legend is made automatically given the variable name and levels.  You do not need to change anything in this code chunk; just run it to see the output 

gf_point(BrainWeight ~ HeadSize, data=braindat, 
         xlab="Head Size (cm^3)", 
         ylab="Brain Weight (grams) of Healthy Adults", 
         color= ~Sex) |>
  gf_lm() |>
  gf_theme(theme_bw())
```


### 1.5. 

To analyze the linear relationships separately for each sex, we need to separate out the cases of each sex from our dataset. Fill in the blanks in the code below to extract out first the male and then the female cases from the original dataset (Hint: you learned this code earlier in the semester -- refer to your appendix/primer if you need a refresher). After you have done so, create the linear model for each. 

```{r}
#| label: build-models-separate
male <- filter(braindat, ____ == "_____")
fema <- filter(braindat, ____ == "_____")

male_lm <- ___(_____ ~ _____, data = male)
fema_lm <- ___(_____ ~ _____, data = fema)
```


### 1.6. 

Write the null and alternative hypotheses for a regression hypothesis of brain weight and head size for *male* healthy adults, both symbolically and verbally. The researcher's hypothesized expectations of the relationship have not changed.

> $H_{0,~male}$:  
> 
> $H_{1,~male}$:  
> 



### 1.7. 

Assess whether the parametric conditions of linear regression have been satisfied, for *only the males*; include any guidelines we should also check for our analyses. Modify the code to create the appropriate graphics. **For just the males**, state the condition, the plot you used as evidence, then whether it has been satisfied. Give a **brief** explanation of the pattern you saw that led to that decision. These are NOT full justifications. (Let your answers in the table wrap if needed; rendering will fix it. Do not add enters, as that will make the table not knit.) 


```{r}
#| label: condition-tests
# condition tests for males



```


*Male Regression*

Condition/Guideline      | Plot            | Satisfied or Violated?  | Pattern (only)
-------------------------|-----------------|-------------------------|-------------------
                         |                 |                         |
                         |                 |                         |
                         |                 |                         |
                         |                 |                         |
                         |                 |                         |



### 1.8. 

Create a summary table for the linear regression for each sex (one summary for males and one summary for females). Using the tables, determine the least squares regression lines for your linear models of each sex.   

<Insert code chunk here>  

> Write linear regression lines here, using markdown symbols  



### 1.9. 

Interpret the R^2^ value of the linear regression for *males*.    

>  Insert answer here  



### 1.10. 

Which sex has a linear regression with a stronger model fit? Provide and compare relevant statistics (one from the model of males and one from the model of females) then justify your answer.  

>  Insert answer here  



### 1.11. 

Determine the 95% confidence interval for the slope of the regression line for *males*. Insert a code to determine the interval, then evaluate the confidence interval for evidence to answer the research question about whether there is a positive linear relationship between brain weight and head size. Be sure to justify your answer in the context of the parameter of interest.   

<Insert code chunk here>  

> Insert answer here  



### 1.12. 

Write a strength of evidence statement for the p-value from your regression analysis of *males*, using a significance level of 0.05.  

>  Insert answer here  



### 1.13. 

TRUE or FALSE (Bold your answer). Both males and females seem to be responding the same in relation to brain weight and head size. (i.e., sex doesn't seem to matter in terms of the direction and strength of the linear association).  



### 1.14. 

TRUE or FALSE (Bold your answer). We could use these equations to predict brain weights beyond the scope of the values of the explanatory variable collected, such as predicting the brain weight for a healthy female adult with a head size of 5000 cm^3^.  


### 1.15. 

The probability that we find no evidence of a linear regression between brain weight and head size, when the truth is there is a linear relationship is a description of what error?  

> Insert answer here  



### 1.16. 

How can you be certain you did not make an error such as described above?  

> Insert answer here  



### 1.17. 

Can you trust the statistical methods used to calculate your inference results (p-value, confidence interval?) for the males? Justify your answer, including what the study has and why or why not that allows you to trust the results.

> Insert answer here  



### 1.18. 

Can we determine if head size **caused** the increase in brain weight in healthy male adults? Explain why or why not with statistical reasoning. 

> Insert answer here  



### 1.20. 

Can we **generalize** our results to all healthy adults? Explain why or why not with statistical reasoning.     

> Insert answer here   




***

### Scenario 2: Agricultural Intensity    


Subsistence farming refers to small agricultural communities that consume most of their crop production. These farms use systems that are extensive, meaning they produce a relatively small amount of crops from a large parcel of land with minimum attention and expense. Naturally, extensive agricultural systems can’t support high population densities because of their relatively low production. Therefore, an increase in population is suspected to be a key factor that forces a shift from extensive to intensive systems of agriculture. The intensity of an agricultural system is defined by the frequency with which a parcel of land is cultivated or is fallow (unplanted) in a crop-fallow cycle. For example a 1 year crop and 0 year fallow cycle would have an agriculture intensity of 100%. While a 1 year crop and 1 year fallow would have an agricultural intensity of 50% and considered less intense than the former example. Researchers are interested in learning about the relationship between agricultural intensity and population density. They gathered data from randomly selected subsistence farms; they did not narrow their focus to farms of any particular crop, but did limit the subsistence farms to ones in tropical environments. In each of these locations, the agricultural intensity, population density, and other variables were recorded. Here population density (in km^2^) is the ratio of the population to the total area available to it for farming. 
You can find the data in `ag-intensity.csv`

 

### 2.1. 

Test the parametric assumptions and guidelines of linear regression for the simple linear regression of agricultural intensity as a function of population density for tropical subsistence farms. Insert the necessary code to create the appropriate graphics. State the condition, and give a complete justification of your assessment of whether it has been met, citing appropriate evidence.  

```{r}
#| label: build-model-ag
ag <- read_csv("ag-intensity.csv", show_col_types = FALSE) # do not change this line of code
names(ag) 

<insert code here>
```

> Condition:  
> Justification:  

> add additional templates as needed




### 2.2. 

Now assess the parametric assumptions and guidelines of linear regression for the transformed linear regression of the log transformed agricultural intensity and population density: `log(AgIntensity)~PopDensity`. The linear model with transformations has been created for you. List each assumption, the graph used to assess it, whether you think the assumption has been satisfied or violated, and a *brief* justification of your answer.  

```{r}
#| label: log-model

ag.log.lm <- lm(log(AgIntensity)~PopDensity, data = ag)

< Insert additional code >
```

Condition                | Plot            | Satisfied or Violated?  | Justification
-------------------------|-----------------|-------------------------|-------------------
                         |                 |                         |
                         |                 |                         |
                         |                 |                         |
                         |                 |                         |
                         |                 |                         |





**Take-home message**: Sometimes, the linear model assumptions are satisfied after taking an appropriate transformation of the explanatory variable and/or the response variable. In the data (agintens.csv), the transformation of the response variable led to (better) satisfying linear model assumptions.

